Equalizer Programming Guide

### 2008-11-13 Developer/ProgrammingGuide/paper.tex
Programming Guide: Prepare for 0.6-rc1 release
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.16 for Equalizer 0.6-RC1\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}
\maketitle
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock}

@@

\thispagestyle{empty}
\section*{Equalizer 0.5 Programming Guide}
\subsection*{Contributors}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

\pagenumbering{arabic}

@@

@@

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
A wall is completely defined by the bottom-left, bottom-right and
top-left coordinates relative to the origin.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eq::net::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
\fig{fFrameSync} illustrates the local frame synchronization.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Blending Slices in 3D-Texture-based
      Volume Rendering}}
\end{wrapfigure}
When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

\cleardoublepage
\thispagestyle{empty}
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
\end{textblock}

>>>>

\date{
  \vspace{15cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.17 for Equalizer 0.6-RC1\\[\medskipamount]
  \today
}

@@

\pagestyle{empty}
\maketitle
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock}

@@

\section*{Equalizer 0.6 Programming Guide}
\subsection*{Contributors}

@@

\clearpage
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
    1.3.17  & Nov 13, 2008 & Added \sref{sLoadBalancing} and \sref{sAdvDPlex}, final touches\\
    \hline
  \end{tabularx}}
\clearpage

\pagenumbering{arabic}
\pagestyle{headings}

@@

\subsection{\label{sLoadBalancing}Load-Balancing}

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-2ex}\includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small Static 2D compound vs 2D
      load-balancing}}
\end{wrapfigure}
While pixel and stereo compounds are naturally load-balanced, 2D, DB and
DPlex compounds might need load-balancing to provide optimal
performance.

Equalizer provides automatic, application-transparent load-balancing for
2D and DB compounds. Using an appropriate load-balancer for these
compounds will adapt the 2D split or database range each frame. The
current load-balancing algorithm is a technology preview and will
improve in future versions.

2D load-balancing gives better framerates than static decomposition in
virtually all cases. DB load-balancing is benefitial for applications
which can not precisely predict the load for their scene data. Volume
rendering is a counter-example, where the data is uniform and a static
DB decomposition typically results in a better performance.

DPlex compounds require a smoothing of the framerate at the destination
channel, otherwise the framerate will become periodically faster and
slower. Using a DPlex load-balancer will smoothen the swap buffer rate
on the destination channel for optimal user experience.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frustra.pdf}
  {\caption{\small\label{fFrustra}Wall and projection parameters}}
\end{wrapfigure}
The frustum description is inherited by the children, therefore the
frustum is typically defined on the top-most compound.

@@

A wall is completely defined by the bottom-left, bottom-right and
top-left coordinates relative to the origin.

@@

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/configInit.pdf}
  {\caption{\small\label{fConfigInit}Config Initialization Sequence}}
\end{wrapfigure}
The \textsf{eqPly::Config} class holds the master versions of the
initialization and frame data. Both objects are registered with the
\textsf{eq::Config}, which is the \textsf{eq::net::Session} used for
rendering. Equalizer takes care of the session setup and exit in
\textsf{Client::choose\-Config} and \textsf{Client::releaseConfig},
respectively.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}\vspace{-5ex}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

The concrete implementation of these window interfaces is responsible
for setting up and de-initializing event handling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eventUML.pdf}
  {\caption{\small\label{fEventUML}UML Diagram for Event Handling Classes}}
\end{wrapfigure}
The glX event handler iterates over all windows of the registered pipe
to find the originating \textsf{GLXWindowIF}, which is then used to
process the event in a fashion similar to AGL. The
\textsf{GLXWindowEvent} passes the \textsf{XEvent} to
\textsf{GLXWindowIF::\-process\-Event}.

@@

\subsubsection{\label{sAdvDPlex}DPlex Compounds}

DPlex decomposition requires multiple frames to be rendered
concurrently. Applications using the thread synchronization models
\textsf{draw\_sync} (the default) or \textsf{local\_sync} have to use
one render client process per GPU to benefit from DPlex task
decomposition.

Non-threaded pipes should not be used for DPlex source and destination
channels. Applications using the \textsf{local\_sync} thread model can
not benefit from DPlex if the application node uses a DPlex source or
destination channel.

Applications using the \textsf{async} thread synchronization model can
fully profit from DPlex using multiple render threads on a multi-GPU
system. For these applications, all render threads run asynchronously
and can render different frames at the same time.

Synchronizing the draw operations between multiple pipe render threads,
and potentially the application thread, breaks DPlex decomposition. At
any given time, only one frame can be rendered from the same
process. The speed-up of DPlex relies however on the capability to
render different frames concurrently.

If one process per GPU is configured, draw-synchronous applications can
scale the performance using DPlex compounds. The processes are not
synchronized with each other, since each process keeps its own version
of the scene data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

\fig{fFrameSync} illustrates the local frame synchronization.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended together to
form a seamless picture. For correct blending, the frames are ordered in
the same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

@@

@@

\cleardoublepage
\pagestyle{empty}
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
\end{textblock}
